---
title: "Binomial risk under absolute value loss: an interesting pattern"
author: "Phillip Nicol"
date: "12/8/2025"
date-modefied: "12/8/2025"
format:
  html:
    toc: true
categories: ["Probability"]
---

If $X \sim \text{Binom}(n,p)$ then the estimator $\hat{p} = X/n$ satisfies $\mathbb{E}(\hat{p} - p)^2 = np(1-p)$, which is quadratic (and concave) in $p$. Instead of using a squared error, we could consider the risk under the absolute value loss:

$$
R(p) = \mathbb{E} | \hat{p} - p|
$$

Below, we plot the risk $R$ for $n = 4,5,10$ and observe an interesting pattern: the risk is not concave and has a cusp at $k/n$. This is much different than the squared error loss!

```{r}
R <- function(n,p) {
  x <- 0:n
  px <- dbinom(x, size=n, prob=p)
  sum(abs(x/n - p)*px)
}
```

```{r, message=FALSE}
library(ggplot2)

n <- c(4,5,10)
p <- seq(0,1,by=0.01)

df <- expand.grid(n=n, p=p)

df$Rp <- apply(df, 1, function(row) R(row[1], row[2]))

df$n <- as.character(df$n)

p <- ggplot(data=df,aes(x=p, y=Rp, color=n)) + geom_line() + 
  theme_bw() + 
  xlab("p") + 
  ylab("Risk")

print(p)
```

The above is the same as the one which was apparently hand-drawn in Blyth (1980).

We can also prove the following result: for $\frac{k-1}{n} \leq p \leq \frac{k}{n}$:

$$
R(p) = 2 \binom{n-1}{k-1} p^k (1-p)^{n-k+1}
$$

This is exercise 1.13 in Lehmann and Casella (1998). We can demonstrate this quickly for $k = 1$:

$$
\begin{align}
R(p) &= p(1-p)^n + \sum_{x=1}^n \left(\frac{k}{n} - p \right) p^x (1-p)^{n-x} \\
&= p(1-p)^n + \mathbb{E}(\hat{p} - p) + (1-p)^n p \\
&= 2(1-p)^n.
\end{align}
$$

Moreover, by differentiating $R(p)$ in this range, we see

$$
R'(p) = 2(1-p)^n - 2np(1-p)^n.
$$

Setting this equal to $0$ and solving for $p$ yields $1/(n+1)$. I.e., the critical point in $[0, 1/n]$ on the above plot occurs at $1/(n+1)$ .

## References 

Blyth, C. R. (1980). Expected absolute error of the usual estimator of the binomial parameter. *The American Statistician*, *34*(3), 155-157.

Lehmann, E. L., & Casella, G. (1998). *Theory of point estimation*. New York, NY: Springer New York.
