---
title: "Ratio of Cauchy: ChatGPT fails this simple question"
author: "Phillip Nicol"
date: "2/8/2025"
date-modefied: "2/8/2025"
format:
  html:
    toc: true
categories: ["Probability"]
---

It is well known that the ratio of two independent standard normals $Z_1/Z_2$ follows a Cauchy distribution. But what about the ratio of two (independent) Cauchy distributions? ChatGPT confidently tells me that it is still a Cauchy distribution. I almost agreed it with it and went on to have a productive day, but got suspicious after the following simulation:

```{r}
set.seed(1)
x <- rcauchy(n=10^4); y <- rcauchy(n=10^4)
r <- x/y

#Quantiles of cauchy 
quantile(x, c(0.025, 0.05, 0.25, 0.5, 0.75, 0.95, 0.975))

#Quantiles of ratio of cauchy
quantile(r, c(0.025, 0.05, 0.25, 0.5, 0.75, 0.95, 0.975))
```

Although the quartiles match up, the ratio appears to have significantly heavier tails.

The density of the ratio $U$, $f_U(u)$, can be computed by the standard approach for (ratio distributions):

```{=tex}
\begin{align}

f_U(u) &= \int_{-\infty}^{\infty} \frac{|v|}{\pi^2 (1 + (uv)^2)(1+v^2)} = \frac{1}{\pi^2} \int_0^{\infty} \frac{1}{\pi^2 (1 + u^2 s)(1+s )ds }

\end{align}
```

Madhav recognized that the integrand has a partial fraction decomposition:

```{=tex}
\begin{equation}
\int_0^{\infty} \frac{1}{\pi^2 (1 + u^2 s)(1+s )} = \frac{1}{\pi^2(u^2-1)} \left( \frac{u^2}{1+u^2 s} - \frac{1}{1+s} \right)
\end{equation}
```
Thus

```{=tex}
\begin{align}
f_U(u) &= \frac{1}{\pi^2 (u^2 - 1)} \left( \lim_{M \to \infty} \int_0^M \frac{u^2}{1+u^2 s} ds - \int_0^M \frac{1}{1+s}ds \right) \\
&= \frac{1}{\pi^2 (u^2 -1)} \lim_{M \to \infty} \left( \ln(1 + u^2 M) - \ln(1+M) \right) \\
&= \frac{\log u^2}{\pi^2 (u^2-1)}
\end{align}
```

which is definitely not a Cauchy distribution. By the way, this is one of the weirdest densities I have ever seen, $f(u) \to \infty$ as $u \to 0$ and $f$ is indeterminate at $u = \pm 1$ (the limit may be evaluated using L'hospital's rule). Here is a plot comparing the Cauchy density (red) to its ratio (black). 

```{r}
x <- seq(-10, 10, by=0.001)
gx <- 1/(pi^2*(1+x^2)) #Standard Cauchy
fx <- (log(x^2))/(pi^2 * (x^2-1)) #Ratio

plot(x,fx,type="l",ylab="Density")
lines(x,gx,type="l", col="red")
```


After deriving this, I found this old paper with the same result:

Ride, P. R. (1965). Distributions of product and quotient of Cauchy variables. *The American Mathematical Monthly*, *72*(3), 303-305.

This means ChatGPT is at least 60 years behind cutting edge statistics!
